Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-18 paper.Rnw
Mallar's comments and fixes based on the reviews

Also of interest to this manuscript are the methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data, of which there are
few. \citet{VanLeemput2009} demonstrate that the applicability of hippocampal
subfield segmentation in T1-weighted images by Bayesian techniques using Markov
random field shape priors learned from 10 manual segmentations. This work,
available as part of the \freesurfer{} package, is limited as the segmentation
omits the tail of the hippocampus and the protocol has yet to be fully
validated. \citet{Yushkevich2009} manually segment hippocampal subfields on
high-resolution (either $0.2mm^3$ isotropic or $0.2mm \times 0.3mm \times 0.2mm$
resolution voxels) T2-weighted MR images acquired from five post-mortem medial
temporal lobe samples. Then, using nonlinear registration guided by shape-based
models of the subfield segmentations, and manually derived hippocampus masks of
the target images, the authors demonstrate accurate parcellation of hippocampal
subfields in clinical 3T T1-weighted MRI volumes. Using \multiatlas{} with bias
correction techniques, \citet{Yushkevich2010} demonstrate a semi-automated
method of subfield segmentation on in vivo focal T2-weighted MR acquisitions of
the temporal lobe. Manual input is only needed to mark divisions between
the head, body and tail of the hippocampus on target images.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 4: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expsubfield{}: \expsubfielddesc{}}
\label{exp:winterburn-xval}
The previous experiment assesses \mb{} performance on whole hippocampus
segmentation. In this experiment, we evaluate \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images at $0.9m^3$-isotropic voxels. We
use a modified leave-one-out cross-validation (LOOCV) design.

@@

\paragraph{Healthy Control Dataset}  %%%%%%%%%%
T1 MR images of 14 subjects were acquired as a part of an ongoing study
at the Centre for Addiction and Mental Health (Table
\ref{tab:WAval-healthy-demographics}). Subjects were known to be free of
neuropsychiatric disorders and gave informed consent. These images were
acquired on a 3T GE Discovery MR 750 system (General Electric, Milwaukee, WI)
using an 8-channel head coil with the enhanced fast gradient recalled echo
3-dimensional acquisition protocol, FGRE-BRAVO, with the following parameters:
$TE/TR/TI = 3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate 0.9mm isotropic voxel
resolution.

@@

\paragraph{Evaluation method}  
Evaluating the agreement of automated hippocampal subfield segmentations with
manual segmentations for T1 images at $0.9mm^3$-isotropic voxels is inherently
ill-defined since there are no manual protocols for segmentation at this
resolution. Instead, we must evaluate the reliability, or {\it precision}, with
which \mb{} produces hippocampal subfields segmentations at this resolution that
correspond in form to the segmentation protocol used by the given
high-resolution atlas library images.

@@

Additionally, by shifting the original manual $0.3mm^3$-isotropic voxel
segmentations by one voxel in the x, y, and z direction and then resampling it
to $0.9mm^3$-isotropic voxels we obtain a simulated manual segmentation having a
small amount of error. We can compare the DSC overlap score of the shifted
labels (relative to the directly resampled labels) with the DSC score of the
\mb{} generated labels in order to evaluate their relevance.

@@

Despite \mb{} achieving segmentation results which are competitive with the rest
of the field (Table \ref{tab:other-methods}), a concern may be raised over the
modest improvement in segmentation agreement observed using \mb{} over
\multiatlas{}, with the same number of atlases (\expadnixval{}). As we have
shown in that same experiment, the benefit in using \mb{} is both an increase in
the overlap agreement and also in the improved consistency of the labelling
regardless of atlas or template choice. Reducing the variability in segmentation
accuracy is an important consideration that few have touched on previously. In
addition, the Monte Carlo cross-validations that we present in \expadnixval{}
and \expfepxval{} are amongst the most stringent performed in the \multiatlas{}
segmentation literature. To the best of our knowledge, with the exception of
\citep{Wang2011}, other groups do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}).  Thus, the
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice.

@@

There are some important differences between our method and these specific
methods. Others have reported the difficulty with mis-registrations in
candidate segmentation (i.e. segmentations generated that are then input in the
voxel-voting procedure \citep{Collins2010}). The work of \citet{Leung2010}
tackles this problem by using an intensity threshold that is estimated
heuristically at the time of segmentation (this work also reports some of the
highest DSC scores for the segmentation of ADNI data). While this method is
effective for the ADNI dataset (which is partially homogenized with respect to
image acquisition and pre-processing), it is unclear if this type of heuristic
is applicable to other datasets. In all cases, these methods require more
atlases than our implementation with the Winterburn atlases.
\citet{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework. In their initial work, \citet{Chupin2009} develop their probabilistic
methodology using a cohort of 8 healthy controls and 15 epilepsy patients, and
then use this method to segment an ADNI sample, with a hierarchical
experimentation protocol. These methods suggest that some post-processing of the
final segmentations would improve accuracy of the segmentation. While that may
be true, there is little consensus regarding how to achieve this.

To the best of our knowledge, no other groups have validated their work using
multiple atlas segmentation protocols, different acquisitions, and disease
populations in order to demonstrate the robustness of their technique. This is
one of the clear strengths of this work. Furthermore, unlike some of the
algorithms mentioned, our implementation does not require retuning for new
populations or datasets as it inherently models the variability of the dataset
through the template library. However it should be noted that the increased
accuracy that follows increasing the number of atlases and templates comes at an
increased computational cost ($O(log(n))$), as previously mentioned in
other work \citep{Heckemann2006}.

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,Yushkevich2010,VanLeemput2009}. Table
\ref{tab:subfield-segmentation-kappa} compares segmentation accuracy from some
of these methods and \mb{}. The overlap DSC scores for \mb{} subfields are
notably lower but a direct comparison of overlap values must be done cautiously.
In the present work, our overlap scores are computed on $0.9mm$-isotropic voxel
resolution images, whereas \citet{Yushkevich2010} uses focal
$0.4\times0.5\times2.0mm$ voxel resolution images, and \citet{VanLeemput2009}
use supersampled $0.380mm$-isotropic voxel resolution images. The larger voxel
images we use necessarily entail a greater change in DSC for each incorrectly
labelled voxel. In addition, our automated segmentations are compared to manual
segmentations resampled from $0.3mm$-isotropic voxel labels; the resampling
process inevitably introduces noise which may lower overlap scores. Lastly, as
our method is aimed specifically at situations when manually produced atlases
are scarce, in our cross validation we are forced to use three rather than all
five of the Winterburn atlases (which, based on our findings with whole
hippocampal segmentation, would have resulted in improved overlap similarity).
Although having more atlases would be ideal in this context, these atlases are
very time consuming to generate \citep{Winterburn2013}. Nevertheless, the
advantage of evaluating \mb{} on standard 3T T1-weighted resolution MR images
with a publically available atlas library is that our results reflect typical
usage scenarios of researchers and clinicians.

@@

In conclusion, we have demonstrated the viability of leveraging a small
number of input atlases to bootstrap a large template library and thereby
improve segmentation reliability when using \multiatlas{} methods. We
demonstrated that this method works robustly over hippocampal definitions,
different disease populations, and different acquisition types. Finally, we also
demonstrate that accurate identification of the hippocampal subfields in
standard 3T T1-weighted images is possible.

>>>>

Also of interest to this manuscript are the methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data, of which there are
few. \citet{VanLeemput2009} demonstrate that the applicability of hippocampal
subfield segmentation in T1-weighted images by Bayesian techniques using Markov
random field shape priors learned from 10 manual segmentations. This work,
available as part of the \freesurfer{} package, is limited as the segmentation
omits the tail of the hippocampus and the protocol has yet to be fully
validated. \citet{Yushkevich2009} manually segment hippocampal subfields on
high-resolution (either $0.2mm$-isotropic or $0.2mm \times 0.3mm \times 0.2mm$
resolution voxels) T2-weighted MR images acquired from five post-mortem medial
temporal lobe samples. Then, using nonlinear registration guided by shape-based
models of the subfield segmentations, and manually derived hippocampus masks of
the target images, the authors demonstrate accurate parcellation of hippocampal
subfields in clinical 3T T1-weighted MRI volumes. Using \multiatlas{} with bias
correction techniques, \citet{Yushkevich2010} demonstrate a semi-automated
method of subfield segmentation on in vivo focal T2-weighted MR acquisitions of
the temporal lobe. Manual input is only needed to mark divisions between
the head, body and tail of the hippocampus on target images.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 4: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expsubfield{}: \expsubfielddesc{}}
\label{exp:winterburn-xval}
The previous experiment assesses \mb{} performance on whole hippocampus
segmentation. In this experiment, we evaluate \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images at $0.9mm$-isotropic voxels. We
use a modified leave-one-out cross-validation (LOOCV) design.

@@

\paragraph{Healthy Control Dataset}  %%%%%%%%%%
T1 MR images of 14 subjects were acquired as a part of an ongoing study
at the Centre for Addiction and Mental Health (Table
\ref{tab:WAval-healthy-demographics}). Subjects were known to be free of
neuropsychiatric disorders and gave informed consent. These images were
acquired on a 3T GE Discovery MR 750 system (General Electric, Milwaukee, WI)
using an 8-channel head coil with the enhanced fast gradient recalled echo
3-dimensional acquisition protocol, FGRE-BRAVO, with the following parameters:
$TE/TR/TI = 3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate $0.9mm$-isotropic voxel
resolution.

@@

\paragraph{Evaluation method}  
Evaluating the agreement of automated hippocampal subfield segmentations with
manual segmentations for T1 images at $0.9mm$-isotropic voxels is inherently
ill-defined since there are no manual protocols for segmentation at this
resolution. Instead, we must evaluate the reliability, or {\it precision}, with
which \mb{} produces hippocampal subfields segmentations at this resolution that
correspond in form to the segmentation protocol used by the given
high-resolution atlas library images.

@@

Additionally, by shifting the original manual $0.3mm$-isotropic voxel
segmentations by one voxel in the x, y, and z direction and then resampling it
to $0.9mm$-isotropic voxels we obtain a simulated manual segmentation having a
small amount of error. We can compare the DSC overlap score of the shifted
labels (relative to the directly resampled labels) with the DSC score of the
\mb{} generated labels in order to evaluate their relevance.

@@

Despite \mb{} achieving segmentation results which are competitive with the rest
of the field (Table \ref{tab:other-methods}), a concern may be raised over the
modest improvement in segmentation agreement observed using \mb{} over
\multiatlas{}, with the same number of atlases (\expadnixval{}). As we have
shown in that same experiment, the benefit in using \mb{} is both an increase in
the overlap agreement and also in the improved consistency of the labelling
regardless of atlas or template choice. Reducing the variability in segmentation
agreement is an important consideration that few have touched on previously. In
addition, the Monte Carlo cross-validations that we present in \expadnixval{}
and \expfepxval{} are amongst the most stringent performed in the \multiatlas{}
segmentation literature. To the best of our knowledge, with the exception of
\citep{Wang2011}, other groups do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}).  Thus, the
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice.

@@

There are some important differences between our method and these specific
methods. Others have reported the difficulty with mis-registrations in
candidate segmentation (i.e. segmentations generated that are then input in the
voxel-voting procedure \citep{Collins2010}). The work of \citet{Leung2010}
tackles this problem by using an intensity threshold that is estimated
heuristically at the time of segmentation (this work also reports some of the
highest DSC scores for the segmentation of ADNI data). While this method is
effective for the ADNI dataset (which is partially homogenized with respect to
image acquisition and pre-processing), it is unclear if this type of heuristic
is applicable to other datasets. In all cases, these methods require more
atlases than our implementation with the Winterburn atlases.
\citet{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework. In their initial work, \citet{Chupin2009} develop their probabilistic
methodology using a cohort of 8 healthy controls and 15 epilepsy patients, and
then use this method to segment an ADNI sample, with a hierarchical
experimentation protocol. These methods suggest that some post-processing of the
final segmentations would improve agreement of the segmentation. While that may
be true, there is little consensus regarding how to achieve this.

To the best of our knowledge, no other groups have validated their work using
multiple atlas segmentation protocols, different acquisitions, and disease
populations in order to demonstrate the robustness of their technique. This is
one of the clear strengths of this work. Furthermore, unlike some of the
algorithms mentioned, our implementation does not require retuning for new
populations or datasets as it inherently models the variability of the dataset
through the template library. However it should be noted that the increased
agreement that follows increasing the number of atlases and templates comes at an
increased computational cost ($O(log(n))$), as previously mentioned in
other work \citep{Heckemann2006}.

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,Yushkevich2010,VanLeemput2009}. Table
\ref{tab:subfield-segmentation-kappa} compares segmentation agreement from some
of these methods and \mb{}. The overlap DSC scores for \mb{} subfields are
notably lower but a direct comparison of overlap values must be done cautiously.
In the present work, our overlap scores are computed on $0.9mm$-isotropic voxel
resolution images, whereas \citet{Yushkevich2010} uses focal
$0.4\times0.5\times2.0mm$ voxel resolution images, and \citet{VanLeemput2009}
use supersampled $0.380mm$-isotropic voxel resolution images. The larger voxel
images we use necessarily entail a greater change in DSC for each incorrectly
labelled voxel. In addition, our automated segmentations are compared to manual
segmentations resampled from $0.3mm$-isotropic voxel labels; the resampling
process inevitably introduces noise which may lower overlap scores. Lastly, as
our method is aimed specifically at situations when manually produced atlases
are scarce, in our cross validation we are forced to use three rather than all
five of the Winterburn atlases (which, based on our findings with whole
hippocampal segmentation, would have resulted in improved overlap similarity).
Although having more atlases would be ideal in this context, these atlases are
very time consuming to generate \citep{Winterburn2013}. Nevertheless, the
advantage of evaluating \mb{} on standard 3T T1-weighted resolution MR images
with a publically available atlas library is that our results reflect typical
usage scenarios of researchers and clinicians.

Experiments 1, 2, and 5 have demonstrated that our algorithm flexibly
accommodates different whole hippocampus manual segmentation methodologies. We
have not explicitly evaluated a subfield definition other than the Winterburn
protocol, and therefore it is possible that using an alternate subfield
definition could improve the reliability of our automated subfield definitions.
For example, established definitions such as those from \citet{Mueller2007}
could be a prime candidate for further exploration. In addition, the
conservative nature of the Mueller definition (labelling of the 5 slices in the
hippocampus body only) would likely further aid in reliability measurement.
However, there are two main logistical problems that we would have to overcome
prior to implementation. The first is that these definitions were developed for
data that is highly anisotropic ($0.4mm \times 0.5mm \times 2mm$), and it is
unclear how our algorithms would deal with such atlases used as input. The
second is that, since these atlases are not publicly available, we would be have
to re-implement the protocol using our atlases. At the present time it is
unclear how we would adapt these protocol to data that we used, where subfield
segmentations are defined on $0.3mm^3$ voxels. However, the impact of subfield
definitions in the context of our work is an important one and should be
considered in subsequent studies.

@@

In conclusion, we have demonstrated the viability of leveraging a small
number of input atlases to bootstrap a large template library and thereby
improve segmentation reliability when using \multiatlas{} methods. We
demonstrated that this method works robustly over hippocampal definitions,
different disease populations, and different acquisition types. Finally, we also
demonstrate that reliable reproduction of hippocampal subfield segmentations in
standard 3T T1-weighted images is possible.
